

# QB Aggressiveness Model

Before we start, my true objective for this project is to expand my R and data analysis skills. I chose football because it's always easier to learn when related to something you are passionate about. So yes, if anyone happens to read this and reconizes it's not the most original idea, I know. That's not the point. The point is to have fun and learn something new. If I can do that while also analyzing football, then it's a win-win. If you happen to find any mistakes in my code or analysis, please let me know. I'm sure there will be plenty and I would love the feedback to improve. Oh, and Go Birds!

## Objective

Measure quarterback aggressiveness using:

- air yards
- tight-window throws
- deep pass rate
- EPA

## Initial Questions

- Which QBs throw deepest?
- Which QBs outperform expected completion probability?
- How stable are aggressiveness metrics year to year?

## Additional Questions to Analyze in the Future

- Which defensive coverages suppress aggressive passing most effectively?
- Do certain defenses force quarterbacks into shorter, lower-EPA throws?
- Can aggressiveness metrics help identify defensive scheme vulnerabilities?


## Methodology

This project evaluates quarterback aggressiveness using NFL play-by-play data from nflverse.

Primary metrics:

- average air yards
- deep throw rate
- EPA per attempt
- completion percentage over expected 

The goal is to separate aggressive decision-making from raw efficiency.



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

```{r load-packages}
library(nflreadr)
library(tidyverse)
library(gt)
library(ggplot2)
```

```{r load-data, cache=TRUE}
#plays <- load_pbp(2015:nflreadr::most_recent_season())
plays <- load_pbp(2025)
```

```{r create-passes, cache=TRUE}
passes <- plays |>
  filter(
    play_type == "pass",
    !is.na(air_yards),
    !is.na(passer_player_name)
  ) |>
  mutate(
    air_yard_bucket = case_when(
      air_yards < 0 ~ "Behind LOS",
      air_yards <= 9 ~ "Short",
      air_yards <= 19 ~ "Intermediate",
      air_yards >= 20 ~ "Deep"
    ),
    air_yard_bucket = factor(
      air_yard_bucket,
      levels = c(
        "Behind LOS",
        "Short",
        "Intermediate",
        "Deep"
      )
    )
  )
```


## League Passing Baselines by Depth

Before evaluating quarterbacks individually we first establish league-wide baseline performance by throw depth.

This helps contextualize:

- how difficult each throw type is
- expected efficiency at each depth
- what qualifies as above-average aggressive passing

```{r depth-summary, cache=TRUE}
depth_summary <- passes |>
  group_by(air_yard_bucket) |>
  summarise(
    attempts = n(),
    avg_epa = mean(epa, na.rm = TRUE),
    completion_pct = mean(complete_pass == 1, na.rm = TRUE),
    avg_yards_gained = mean(yards_gained, na.rm = TRUE) 
  ) |>
  arrange(air_yard_bucket)
```
```{r show-depth-summary}
depth_summary |>
  gt()
```

#### Interesting Observations
 
 - Deeper throws have the highest epa but lowest competion percentage. 
 - Throws behind the line of scrimmage are the only bucket with negative average EPA. 


## QB Aggressiveness Leaderboard

This leaderboard ranks quarterbacks by average air yards per attempt to identify which passers push the ball downfield most aggressively. Additional efficiency metrics help distinguish productive aggressiveness from inefficient high-risk passing.

```{r create-qb-summary, cache=TRUE}
qb_summary <- passes |>
  group_by(passer_player_name) |>
  summarise(
    attempts = n(),
    avg_air_yards = mean(air_yards, na.rm = TRUE),
    deep_pass_rate = mean(air_yards >= 20, na.rm = TRUE),
    epa_per_play = mean(epa, na.rm = TRUE),
    completion_pct = mean(complete_pass == 1, na.rm = TRUE)
  ) |>
  filter(attempts >= 100) |>
  arrange(desc(avg_air_yards))
```

```{r show-qb-aggressiveness-leaderboard}
qb_summary |>
  select(
    passer_player_name,
    attempts,
    avg_air_yards,
    deep_pass_rate,
    epa_per_play,
    completion_pct
  ) |>
  arrange(desc(avg_air_yards)) |>
  slice_head(n = 15) |>
  gt()
```


## Best Deep Passers

These quarterbacks generated the most EPA above league average on deep throws. Minimum 25 deep attempts to ensure a stable sample size.

```{r qb-vs-league, cache=TRUE}
# Summarize each player by depth bucket
qb_vs_league <- passes |>
  group_by(passer_player_name, air_yard_bucket) |>
  summarise(
    attempts = n(),
    avg_epa = mean(epa, na.rm = TRUE),
    completion_pct = mean(complete_pass == 1, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(attempts >= 25)
```




```{r qb-vs-league-delta, cache=TRUE}
# Join QB results to league baselines and calculate deltas
qb_vs_league <- qb_vs_league |>
  left_join(
    depth_summary |>
      select(
        air_yard_bucket,
        league_epa = avg_epa,
        league_completion_pct = completion_pct
      ),
    by = "air_yard_bucket"
  ) |>
  mutate(
    epa_over_league = avg_epa - league_epa,
    completion_over_league =
      completion_pct - league_completion_pct
  )
```




```{r show-best-deep-passers}
qb_vs_league |>
  filter(
    air_yard_bucket == "Deep",
    attempts >= 25
  ) |>
  arrange(desc(epa_over_league)) |>
  slice_head(n = 15) |>
  gt()
```

## QB Agressiveness vs Efficiency
```{r show-aggressiveness-vs-efficiency, fig.width=12, fig.height=8}
qb_summary |>
  ggplot(aes(
    x = epa_per_play,
    y = avg_air_yards,
  )) +
  geom_point() +
  geom_text(
    aes(label = passer_player_name),
    hjust = -0.1,
    size = 3
  ) +
  geom_vline(
    xintercept = mean(qb_summary$epa_per_play),
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = mean(qb_summary$avg_air_yards),
    linetype = "dashed"
  ) +
  labs(
    title = "QB Aggressiveness vs Efficiency",
    x = "EPA Per Play",
    y = "Average Air Yards"
  )
```